% ============================================================================
% EXERCISE B: NITROGEN RESPONSE FACTORIAL
% ============================================================================
% Duration: 90 minutes (hands-on; Module 4 support)
% Learning Goals: N response curves; AEN calculation; cost-benefit analysis
% Prerequisites: Module 3, Module 4

\chapter{Exercise B: Nitrogen Response Factorial}
\label{ch:exercise-b}

\section{Overview}

\textbf{Objective:} Run four N rate treatments and analyze yield response to determine optimal N rate and calculate agronomic efficiency.

\textbf{Files:} 
\begin{itemize}
  \item \filename{Exercise\_B\_N0.apsimx} (0 kg/ha N)
  \item \filename{Exercise\_B\_N80.apsimx} (80 kg/ha N)
  \item \filename{Exercise\_B\_N160.apsimx} (160 kg/ha N)
  \item \filename{Exercise\_B\_N240.apsimx} (240 kg/ha N)
\end{itemize}

\textbf{Time:} 90 minutes (run all 4 simulations; ~10 min per run)

\textbf{Deliverable:} Excel workbook with N response curve, AEN calculations, cost-benefit analysis.

\section{Simulation Setup}

\subsection{Configuration}

\begin{table}[H]
\centering
\caption{Exercise B Simulation Parameters}
\label{tab:exercise-b-setup}
\begin{tabularx}{\textwidth}{l|X}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Location & Dalby, Queensland \\
Climate & 1949--1950 (single growing season) \\
Weather file & AU\_Dalby.met \\
Soil & Norwin Vertosol \\
Crop & Wheat (Hartog cultivar) \\
Sowing date & May 15 (variable rule: ESW $>$ 100 mm + rain threshold) \\
Fertiliser rate & 0, 80, 160, 240 kg/ha NO\textsubscript{3}-N (applied at sowing) \\
Reported variables & Clock.Today, Wheat.LAI, Wheat.Phenology.Zadok.Stage, Wheat.Grain.Total.Wt, Wheat.Grain.Protein, Wheat.AboveGround.N \\
\bottomrule
\end{tabularx}
\end{table}

\section{Task 1: Run All Four Simulations}

\begin{enumerate}
  \item Open Exercise\_B\_N0.apsimx; run (export to \filename{Exercise\_B\_N0\_Results.xlsx})
  \item Open Exercise\_B\_N80.apsimx; run (export to \filename{Exercise\_B\_N80\_Results.xlsx})
  \item Open Exercise\_B\_N160.apsimx; run (export to \filename{Exercise\_B\_N160\_Results.xlsx})
  \item Open Exercise\_B\_N240.apsimx; run (export to \filename{Exercise\_B\_N240\_Results.xlsx})
  
  \textit{Note:} Each simulation should take $<$ 5 minutes
\end{enumerate}

\section{Task 2: Extract Grain Yield}

\subsection{Approach}

\begin{itemize}
  \item For each .xlsx file, find the row with maximum Wheat.Grain.Total.Wt
  \item This represents harvest time (physiological maturity)
  \item Record as final yield for that N rate
  \item Create summary table: N rate vs. Final Yield
\end{itemize}

\begin{table}[H]
\centering
\caption{Expected N Response Pattern}
\label{tab:n-response}
\begin{tabularx}{\textwidth}{r|r|r}
\toprule
\textbf{N Rate (kg/ha)} & \textbf{Yield (kg/ha)} & \textbf{AEN (kg/kg)} \\
\midrule
0 & 2000 & -- \\
80 & 3500 & (3500--2000)/80 = 18.75 \\
160 & 4600 & (4600--2000)/160 = 16.25 \\
240 & 5200 & (5200--2000)/240 = 13.33 \\
\bottomrule
\end{tabularx}
\end{table}

\textit{(Exact numbers will vary; this is illustrative)}

\section{Task 3: Calculate Agronomic Efficiency of N (AEN)}

\[
\text{AEN}_{i} = \frac{\text{Yield}_{i} - \text{Yield}_{0 \text{ kg/ha}}}{N \text{ rate}_{i}} \quad [\text{kg grain / kg N}]
\]

Example calculation:
\begin{itemize}
  \item Yield(N=0) = 2000 kg/ha
  \item Yield(N=80) = 3500 kg/ha
  \item AEN(80) = (3500 -- 2000) / 80 = 18.75 kg grain / kg N
\end{itemize}

\textbf{Observations:}
\begin{itemize}
  \item AEN is highest at low N rates (diminishing returns)
  \item AEN decreases as N rate increases (approaching physiological limit)
  \item AEN reflects N use efficiency; typical wheat: 10--20 kg/kg
\end{itemize}

\section{Task 4: Plot N Response Curve}

\begin{itemize}
  \item X-axis: N rate (0, 80, 160, 240 kg/ha)
  \item Y-axis: Yield (kg/ha)
  \item Plot points for each N rate
  \item Add polynomial trendline (degree 2 is typical)
  \item Title: ``Nitrogen Response Curve: Wheat''
\end{itemize}

\textbf{Interpretation:} Where is the curve flat (diminishing returns)? At what N rate does yield plateau?

\section{Task 5: Cost-Benefit Analysis}

\subsection{Profit Calculation}

\[
\text{Profit} = (\text{Yield} \times P_{\text{grain}}) - (\text{N rate} \times P_{\text{N}})
\]

\subsection{Assumptions}

\begin{itemize}
  \item Grain price: \$400/tonne (adjust based on current market)
  \item N cost: \$1.50/kg (adjust based on fertiliser price)
\end{itemize}

\subsection{Example Calculation}

For N = 160 kg/ha:
\begin{itemize}
  \item Yield = 4600 kg/ha = 4.6 tonnes/ha
  \item Revenue = 4.6 × \$400 = \$1840/ha
  \item N cost = 160 × \$1.50 = \$240/ha
  \item Profit = \$1840 -- \$240 = \$1600/ha
\end{itemize}

\subsection{Create Profit Table}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{r|r|r|r|r}
\toprule
\textbf{N (kg/ha)} & \textbf{Yield (t/ha)} & \textbf{Revenue} & \textbf{N Cost} & \textbf{Profit} \\
\midrule
0 & -- & -- & \$0 & -- \\
80 & -- & -- & \$120 & -- \\
160 & -- & -- & \$240 & -- \\
240 & -- & -- & \$360 & -- \\
\bottomrule
\end{tabularx}
\end{table}

\section{Task 6: Analysis Questions}

\begin{enumerate}
  \item What is the yield response to N? (How much yield increases per kg N?)
  
  \item At what N rate does AEN drop to 10 kg/kg? (Where do diminishing returns become pronounced?)
  
  \item What is the profit-maximizing N rate (assuming prices: \$400/tonne grain, \$1.50/kg N)?
  
  \item If grain price fell to \$300/tonne, would you increase or decrease N rate? Why?
  
  \item If N cost rose to \$2.50/kg, how would profit be affected?
\end{enumerate}

\section{Task 7: Grain Quality}

\subsection{Protein Analysis}

\begin{itemize}
  \item Extract Wheat.Grain.Protein from each simulation at harvest
  \item Plot: N rate vs. Protein content
  \item Interpretation: Does higher N increase grain protein? By how much?
\end{itemize}

\textit{Expected finding:} Grain protein increases with N rate (premium prices for high-protein wheat).

\section{Deliverable}

Submit \filename{Exercise\_B\_Report.xlsx} or \filename{Exercise\_B\_Report.docx}:

\begin{itemize}
  \item $\checkmark$ Yield summary table (N rate vs. Yield)
  \item $\checkmark$ AEN calculations (table with AEN for each N rate)
  \item $\checkmark$ N response curve plot
  \item $\checkmark$ Profit analysis table + chart
  \item $\checkmark$ Protein vs. N rate plot
  \item $\checkmark$ Written answers to 7 analysis questions
  \item $\checkmark$ Conclusion: Your recommendation for optimal N rate and justification
\end{itemize}

% ============================================================================
% END EXERCISE B
% ============================================================================
