% ============================================================================
% EXERCISE D: 20-YEAR VARIABILITY ANALYSIS
% ============================================================================
% Duration: 120 minutes (hands-on; Module 6 support)
% Learning Goals: Multi-year yield variability; risk assessment; robust decisions
% Prerequisites: Module 3, Module 6

\chapter{Exercise D: 20-Year Variability Analysis}
\label{ch:exercise-d}

\section{Overview}

\textbf{Objective:} Run a 20-year continuous simulation and analyze yield distribution, identify dry/wet years, assess management robustness.

\textbf{File:} \filename{Exercise\_D\_20Year\_Variability.apsimx}

\textbf{Time:} 120 minutes (includes data processing and visualization)

\textbf{Deliverable:} Comprehensive analysis of yield variability and risk metrics.

\section{Simulation Setup}

\subsection{Configuration}

\begin{table}[H]
\centering
\caption{Exercise D Simulation Parameters}
\label{tab:exercise-d-setup}
\begin{tabularx}{\textwidth}{l|X}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Location & Dalby, Queensland \\
Climate & 20 consecutive years (e.g., 1950--1969) \\
Weather file & AU\_Dalby.met \\
Soil & Norwin Vertosol \\
Crop & Wheat (Hartog cultivar) \\
Management & Fixed: Sow May 1--July 10 (ESW $>$ 100 mm); N = 160 kg/ha \\
Report variables & Clock.Today, Wheat.Grain.Total.Wt, Weather.Rainfall, Soil.ESW \\
\bottomrule
\end{tabularx}
\end{table}

\section{Task 1: Run Simulation}

\begin{enumerate}
  \item Open \filename{Exercise\_D\_20Year\_Variability.apsimx}
  \item Verify Clock: start 1950-01-01, end 1969-12-31 (20 years)
  \item Run simulation ($\sim$ 30 seconds)
  \item Export to \filename{Exercise\_D\_Results.xlsx}
  \item Verify: $\approx$ 7300 daily records (365 × 20 years)
\end{enumerate}

\section{Task 2: Extract Annual Yields}

\subsection{Approach}

\begin{enumerate}
  \item Create helper column: Year (extract from date)
  \item Create helper column: Season Year (e.g., May 1, 1950 = year 1950 season; previous May 1 thru Apr 30)
  \item For each season year, find maximum Wheat.Grain.Total.Wt (represents harvest)
  \item Create summary table: Season Year vs. Final Yield
\end{enumerate}

\subsection{Result Example}

\begin{table}[H]
\centering
\caption{Annual Yield Summary (Illustrative)}
\label{tab:annual-yield}
\begin{tabularx}{\textwidth}{r|r|r|r}
\toprule
\textbf{Season Year} & \textbf{Annual Rainfall (mm)} & \textbf{Final Yield (kg/ha)} & \textbf{ESW at Sowing (mm)} \\
\midrule
1950 & 450 & 3800 & 125 \\
1951 & 380 & 2100 & 95 \\
\textit{(wet year)} & \textit{(high)} & \textit{(high)} & -- \\
\textit{(dry year)} & \textit{(low)} & \textit{(low)} & -- \\
... & ... & ... & ... \\
\bottomrule
\end{tabularx}
\end{table}

\section{Task 3: Calculate Risk Metrics}

\subsection{Basic Statistics}

\begin{itemize}
  \item Mean yield: $\bar{Y} = \frac{1}{n} \sum Y_i$
  \item Std dev: $\sigma = \sqrt{\frac{1}{n-1} \sum (Y_i - \bar{Y})^2}$
  \item Min/Max: Identify driest and wettest years
  \item Coefficient of variation: $\text{CV} = \sigma / \bar{Y}$
\end{itemize}

\subsection{Risk Metrics}

\[
P(\text{Yield} < 1500 \text{ kg/ha}) = \frac{\text{\# years with Yield} < 1500}{20 \text{ years}}
\]

\textit{Interpretation:} If P = 0.25, then 1 in 4 years (25\%) has yield below 1500 kg/ha; risky threshold.

\begin{keyconceptbox}
\textbf{Robustness Definition:} A management practice is ``robust'' if it maintains yield above a threshold (e.g., 2000 kg/ha) in at least 80\% of years. High-risk scenarios fail in $>$ 20\% of years.
\end{keyconceptbox}

\section{Task 4: Visualize Yield Distribution}

\subsection{Plot 1: Annual Yield Time Series}

\begin{itemize}
  \item X-axis: Year (1950--1969)
  \item Y-axis: Yield (kg/ha)
  \item Add horizontal lines: Mean, Mean±1SD (shaded band)
  \item Title: ``Annual Wheat Yield: 20-Year Variability''
\end{itemize}

\subsection{Plot 2: Histogram of Yields}

\begin{itemize}
  \item X-axis: Yield classes (e.g., 1000--1500, 1500--2000, etc.)
  \item Y-axis: Frequency (number of years)
  \item Overlay: Normal distribution curve
  \item Title: ``Yield Distribution''
\end{itemize}

\subsection{Plot 3: Rainfall vs. Yield Scatter}

\begin{itemize}
  \item X-axis: Annual rainfall
  \item Y-axis: Annual yield
  \item Add regression line
  \item Title: ``Rainfall vs. Yield Relationship''
\end{itemize}

\section{Task 5: Identify Dry and Wet Years}

\subsection{Classification}

Define dry/wet years using rainfall or yield:
\begin{itemize}
  \item Dry: Yield $<$ (Mean -- 1 SD) or Rainfall $<$ 25th percentile
  \item Average: (Mean -- 0.5 SD) to (Mean + 0.5 SD)
  \item Wet: Yield $>$ (Mean + 1 SD) or Rainfall $>$ 75th percentile
\end{itemize}

\subsection{Analysis}

\begin{enumerate}
  \item Identify specific dry years (e.g., 1952, 1957)
  \item Identify specific wet years (e.g., 1955, 1960)
  \item Compare phenology, grain number, protein for dry vs. wet years
  \item Hypothesize: Why does 1952 fail while 1951 doesn't?
\end{enumerate}

\section{Task 6: Analysis Questions}

\begin{enumerate}
  \item What is the average yield? Coefficient of variation? Interpretation in terms of farmer risk?
  
  \item In how many years does yield exceed 3000 kg/ha? (Probability of good harvest)
  
  \item In how many years does yield fall below 1500 kg/ha? (Risk threshold)
  
  \item Which weather variable (rainfall, temperature, solar radiation) most affects yield in this region?
  
  \item If a farmer needs minimum 2000 kg/ha to be profitable, what is their risk of failure? Acceptable?
  
  \item Propose an adaptation: Should farmers shift sowing date (earlier? later?) to reduce risk? Or increase N? Or plant a different variety? Justify using your data.
\end{enumerate}

\section{Task 7: Scenario Analysis (Extension)}

\subsection{Scenario A: Increased N (240 kg/ha)}

Re-run simulation with higher N. Does it reduce year-to-year variability? Why or why not?

\subsection{Scenario B: Earlier Sowing}

Shift sowing window to April--June. How does yield variability change?

\subsection{Scenario C: Climate Change}

Modify weather data (e.g., +2°C, --10\% rainfall). Run 20-year simulation. Compare yield distribution.

\section{Deliverable}

Submit \filename{Exercise\_D\_Report.xlsx} or \filename{Exercise\_D\_Report.docx}:

\begin{itemize}
  \item $\checkmark$ Annual yield summary table (20 years)
  \item $\checkmark$ Risk metrics (mean, SD, CV, P(Yield $<$ threshold))
  \item $\checkmark$ 3+ visualizations (time series, histogram, scatter)
  \item $\checkmark$ Dry vs. wet year comparison (phenology, yields)
  \item $\checkmark$ Written answers to 7 analysis questions
  \item $\checkmark$ Recommendation: Is the current management robust? What adaptation would you propose?
\end{itemize}

% ============================================================================
% END EXERCISE D
% ============================================================================
