% ============================================================================
% MAIN.TEX: SCIE3314 APSIM Tutorial
% ============================================================================
% Root document that compiles all modules, exercises, and appendices
% 
% COMPILATION:
%   pdflatex main.tex
%   bibtex main
%   pdflatex main.tex
%   pdflatex main.tex
%
% Or use latexmk: latexmk -pdf main.tex

\input{preamble}

% ============================================================================
% CUSTOM TITLE PAGE FORMATTING
% ============================================================================

\makeatletter
\renewcommand{\maketitle}{%
  \thispagestyle{empty}
  \begin{tikzpicture}[remember picture, overlay]
    % Blue header bar
    \node[anchor=north west, fill=UWABlue, minimum width=\paperwidth, minimum height=6.5cm] 
      (header) at (current page.north west) {};
    % Gold accent stripe
    \node[anchor=south west, fill=UWAGold, minimum width=\paperwidth, minimum height=0.55cm] 
      at (header.south west) {};
    % Main title
    \node[anchor=west, text=white, font=\Huge\bfseries, text width=14cm, inner sep=0pt, align=left]
      at ([xshift=2cm, yshift=-2.8cm]current page.north west) {\@title};
    % Author/Institution
    \node[anchor=west, text=white, font=\Large, inner sep=0pt, align=left]
      at ([xshift=2cm, yshift=-4.6cm]current page.north west) {\@author};
    % Date
    \node[anchor=west, text=white, font=\normalsize, inner sep=0pt, align=left]
      at ([xshift=2cm, yshift=-5.5cm]current page.north west) {\@date};
  \end{tikzpicture}
  \vspace{5.0cm}
}
\makeatother

\title{%
  SCIE3314: Agricultural Production \\[0.2em]%
  Systems Simulator (APSIM) \\[0.3em]%
  \large Introductory Tutorial \\
  \normalsize Modular Design for Crop Modelling and Decision Support
}

\author{
  School of Agriculture and Environment \\
  The University of Western Australia \\
  \texttt{Gustavo.Alckmin@uwa.edu.au}
}

\date{May 2026 \\ Version 1.0}

\begin{document}

% ============================================================================
% FRONT MATTER
% ============================================================================

\frontmatter

% Title page
\maketitle

% ============================================================================
% ACKNOWLEDGMENTS
% ============================================================================

\chapter*{Acknowledgments}

This tutorial was developed with reference to:
\begin{itemize}
  \item APSIM Next Gen 2025.6.7774.0 documentation
  \item APSOIL soil database (Australian soils)
  \item Weather files from BOM/SILO networks
% \item Pedagogic design based on Bloom's taxonomy and constructivist learning principles
\end{itemize}

We acknowledge the APSIM Initiative (CSIRO, QAAFI, CIMMYT, ICRISAT, Michigan State University, etc) for developing and maintaining APSIM as a public good.

% ============================================================================
% TABLE OF CONTENTS, FIGURES, TABLES
% ============================================================================

\tableofcontents
\listoffigures
\listoftables

% ============================================================================
% PREFACE
% ============================================================================

\chapter*{Preface}

This tutorial introduces APSIM (Agricultural Production Systems Simulator), a process-based crop modelling platform widely used in research, extension, and operational decision support.

\begin{keyconceptbox}
\textbf{Learning Outcomes:}
After completing this tutorial, students will be able to:
\begin{enumerate}
  \item Understand the structure and daily sequence of APSIM simulations
  \item Configure and run crop simulations for decision analysis
  \item Interpret model outputs and assess crop responses to management
  \item Apply APSIM to explore agronomic questions (N fertilisation, phenology, variability)
  \item Troubleshoot common errors and validate simulation logic
\end{enumerate}
\end{keyconceptbox}

\section*{How to Use This Tutorial}

\textbf{For Instructors:} Each module (0--7) is designed as a 30--90 minute class session or self-study unit. Four integrated exercises (A--D) scaffold hands-on simulation work, progressing from simple water balance to multi-year variability analysis. All exercises use near-solved .apsimx files (components pre-configured; students modify parameters).

\textbf{For Students:} Start with Module 0 (Why APSIM?), then progress sequentially through Modules 1--7. Complete Exercises A--D in parallel or after each module block. Each module includes key concepts, worked examples, and links to reference entities in the LLM Wiki (\textit{see \cref{sec:wiki-resources}}).

\textbf{Prerequisites:} Basic agronomy (crop growth stages, soil water, nitrogen cycling). No programming required; JavaScript Manager rules are provided as templates.

% ============================================================================
% MAIN MATTER
% ============================================================================

\mainmatter

% ============================================================================
% MODULE 0: WHY APSIM?
% ============================================================================

\input{modules/module_0_why_apsim}

% ============================================================================
% MODULE 1: STRUCTURE
% ============================================================================

\input{modules/module_1_structure}

% ============================================================================
% MODULE 2: FIRST SIMULATION
% ============================================================================

\input{modules/module_2_first_simulation}

% ============================================================================
% MODULE 3: OUTPUT ANALYSIS
% ============================================================================

\input{modules/module_3_output_analysis}

% ============================================================================
% MODULE 4: NITROGEN FERTILISATION
% ============================================================================

\input{modules/module_4_nitrogen_fertilisation}

% ============================================================================
% MODULE 5: PHENOLOGY & THERMAL TIME
% ============================================================================

\input{modules/module_5_phenology_thermal_time}

% ============================================================================
% MODULE 6: LONG-TERM VARIABILITY
% ============================================================================

\input{modules/module_6_long_term_variability}

% ============================================================================
% MODULE 7: CLIMATE SCENARIOS
% ============================================================================

\input{modules/module_7_climate_scenarios}

% ============================================================================
% EXERCISES
% ============================================================================

\appendix

\part*{Integrated Exercises}
\addcontentsline{toc}{part}{Integrated Exercises}

% Exercise A
\input{exercises/exercise_a_fallow_water_balance}

% Exercise B
\input{exercises/exercise_b_nitrogen_response}

% Exercise C
\input{exercises/exercise_c_sowing_date}

% Exercise D
\input{exercises/exercise_d_20_year_variability}

% ============================================================================
% APPENDICES
% ============================================================================

\input{appendix/appendix_a_components_reference}

\input{appendix/appendix_b_troubleshooting}

% ============================================================================
% BACK MATTER
% ============================================================================

\backmatter

% ============================================================================
% REFERENCES
% ============================================================================

\chapter{References}

\begin{thebibliography}{99}

\bibitem{APSIM2025}
  APSIM Initiative (2025). \textit{APSIM Next Gen Documentation}.
  \url{https://www.apsim.info}

\bibitem{Holzworth2014}
  Holzworth, D.P., Huth, N.I., deVoil, P.G., et al.\ (2014).
  APSIM --- Evolution towards a new generation of agricultural systems simulation.
  \textit{Environmental Modelling \& Software}, 62, 327--350.
  \url{https://doi.org/10.1016/j.envsoft.2014.07.009}

\bibitem{Keating1998}
  Keating, B.A., Carberry, P.S., Hammer, G.L., et al.\ (1998).
  An overview of APSIM, a model designed for farming systems simulation.
  \textit{European Journal of Agronomy}, 18, 267--288.

\end{thebibliography}

% ============================================================================
% WIKI RESOURCES
% ============================================================================

\chapter{LLM Wiki Resources}
\label{sec:wiki-resources}

Throughout this tutorial, we reference concepts stored in the SCIE3314 APSIM Wiki:

\begin{description}
  \item[\apsimterm{Entities}] Definition pages for APSIM components (Clock, Weather, Soil, Manager, Report, DataStore, Wheat, etc.) with misconceptions and troubleshooting.
  \item[\apsimterm{Concepts}] Pedagogic synthesis pages (Scaffolding, Near-Solved Design, Water Stress, Phenology-Thermal-Time Link, NUE, Module Dependencies).
  \item[\apsimterm{FAQ}] Diagnostic pages for common errors and decisions.
  \item[\apsimterm{Exercises}] Step-by-step guides for Exercises A--D with learning objectives and analysis prompts.
\end{description}

Wiki location: \url{https://github.com/00113844/SCIE3314_APSIM_Tutorial/tree/main/SCIE3314_APSIM_Wiki}

% ============================================================================
% FURTHER RESOURCES
% ============================================================================

\chapter{Further Resources}

\begin{itemize}
  \item \textbf{APSIM Next Gen documentation and tutorials:}\\
        \url{https://apsimnextgeneration.netlify.app/user_tutorials/}
  \item \textbf{APSIM community forum and support:}\\
        \url{https://www.apsim.info/}
  \item \textbf{ApsimX source code (GitHub):}\\
        \url{https://github.com/APSIMInitiative/ApsimX}
  \item Holzworth, D.P.\ et al.\ (2014). APSIM --- Evolution towards a new generation
        of agricultural systems simulation.
        \textit{Environmental Modelling \& Software}, 62, 327--350.
        \url{https://doi.org/10.1016/j.envsoft.2014.07.009}
\end{itemize}

% ============================================================================
% INSTRUCTOR NOTES
% ============================================================================

\chapter{Instructor Notes}

\begin{itemize}
  \item Partial \texttt{.apsimx} exercise files (A--D) should be distributed
        via LMS \emph{before} the first session.
  \item All four exercises require APSIM Next Gen $\geq$ version 2023.2.7164.0.
  \item Expected run times: Exercises A, B, C $<$30 seconds each;
        Exercise D $\approx$1--2 minutes.
  \item Exercise C requires the \texttt{Experiment} node --- ensure students
        have completed Module 1 (component reference) before starting.
  \item Exercises B and D are suitable for individual assessment: the data tables
        in each exercise can be used as submission items.
  \item Required files: \texttt{wheat\_fallow\_partial.apsimx},
        \texttt{wheat\_Nfert\_partial.apsimx},
        \texttt{wheat\_sowingdate\_partial.apsimx},
        \texttt{wheat\_whatif\_partial.apsimx} (provided via LMS).
\end{itemize}

\end{document}

% ============================================================================
% END MAIN.TEX
% ============================================================================
