% ============================================================================
% APPENDIX B: TROUBLESHOOTING COMMON ERRORS
% ============================================================================

\chapter{Troubleshooting Common Errors}
\label{app:troubleshooting}

This appendix provides diagnostic decision trees for common APSIM problems.

\section{Crop Won't Sow}

\subsection{Diagnostic Flowchart}

\begin{enumerate}
  \item \textbf{Q: Does simulation run without errors?}
    \begin{itemize}
      \item YES $\to$ Go to 2
      \item NO $\to$ Check file format, component structure; see APSIM documentation
    \end{itemize}
  
  \item \textbf{Q: Check final output: Is Zadok stage always 0 (not sown)?}
    \begin{itemize}
      \item YES $\to$ Crop is not sowing; go to 3
      \item NO $\to$ Crop did sow at some point; issue is different
    \end{itemize}
  
  \item \textbf{Q: Is there a Manager rule for sowing?}
    \begin{itemize}
      \item YES $\to$ Go to 4
      \item NO $\to$ Add Manager rule (see Module 1); crop won't sow without rule
    \end{itemize}
  
  \item \textbf{Q: Check Manager parameters: Is ESW threshold too high?}
    \begin{itemize}
      \item Check: \variable{MinESW} parameter. If MaxESW in simulation $<$ MinESW, crop never meets threshold
      \item Solution: Reduce MinESW or increase initial soil water
    \end{itemize}
  
  \item \textbf{Q: Check Manager parameters: Is sowing window correct?}
    \begin{itemize}
      \item Check: \variable{StartDate}, \variable{EndDate}. Are they in correct format (d-mmm)?
      \item Example: ``1-may'' (correct), ``May 1'' (incorrect)
      \item Solution: Use d-mmm format
    \end{itemize}
  
  \item \textbf{Q: Is rainfall threshold met?}
    \begin{itemize}
      \item Check: Weather file data. Is there a 7-day period with rainfall $>$ 25 mm during sowing window?
      \item Solution: If region is always dry, reduce rainfall threshold or choose wetter year from weather file
    \end{itemize}
\end{enumerate}

\section{Unrealistic Yield}

\subsection{Yield Too Low}

\begin{warningbox}
\textbf{Possible causes:}
\begin{enumerate}
  \item Crop sown too late (short season, incomplete grain fill)
  \item Insufficient N (N stress throughout season)
  \item Water stress (low rainfall or high drainage)
  \item Crop encounters frost during booting/anthesis
\end{enumerate}
\end{warningbox}

\textbf{Diagnostic:}
\begin{itemize}
  \item Plot Zadok stage vs. date; check if crop reaches grain fill (stages 70--89)
  \item Plot [Soil].ESW vs. date; check for critical stress periods (booting, grain fill)
  \item Check Manager: Is sowing happening at right time?
  \item Check N rate: Is it sufficient (compare N=0 vs. N=160)?
\end{itemize}

\subsection{Yield Too High}

\begin{warningbox}
\textbf{Possible causes:}
\begin{enumerate}
  \item Initial conditions unrealistic (soil too wet, too much initial N)
  \item Weather file synthetic or biased (unrealistically favorable)
  \item Report variables wrong (grain mass not realistic for cultivar/location)
\end{enumerate}
\end{warningbox}

\textbf{Diagnostic:}
\begin{itemize}
  \item Compare simulated yield to local farm data / research records
  \item Check initial soil water (\variable{InitialPAWmm}) and initial N pools
  \item Verify weather file source and quality
\end{itemize}

\section{No Output or Empty Report}

\subsection{DataStore Empty}

\begin{warningbox}
\textbf{Possible causes:}
\begin{enumerate}
  \item Report component missing or disabled
  \item Report variables list is empty
  \item Simulation didn't run (errors in setup)
\end{enumerate}
\end{warningbox}

\textbf{Diagnostic:}
\begin{itemize}
  \item In APSIM GUI, click Report in tree; check Properties panel
  \item Verify: VariableNames array is not empty
  \item Verify: Report is Enabled (checkbox ticked)
  \item Re-run simulation; check output window for errors
\end{itemize}

\subsection{Report Variables Have Missing Values}

\begin{warningbox}
\textbf{Possible causes:}
\begin{enumerate}
  \item Variable path is incorrect (typo in [Component].Variable naming)
  \item Component not initialized when Report tries to access it (e.g., [Wheat].LAI before crop sown)
  \item Component is disabled
\end{enumerate}
\end{warningbox}

\textbf{Diagnostic:}
\begin{itemize}
  \item Verify variable names match APSIM conventions (\cref{sec:var-naming})
  \item Use APSIM GUI to auto-generate variable names (reduces typos)
  \item Check if missing values coincide with off-season (e.g., wheat LAI=0 when crop not alive)
\end{itemize}

\section{File Path Errors}

\begin{warningbox}
\textbf{Error message:} ``Cannot find weather file AU\_Northam.met''
\end{warningbox}

\textbf{Diagnosis:}
\begin{itemize}
  \item Check Weather component: FileName property uses \variable{\%root\%} variable
  \item \variable{\%root\%} = APSIM installation directory
  \item Example: \texttt{C:/Program Files/APSIM2025/Examples/WeatherFiles/AU\_Northam.met}
\end{itemize}

\textbf{Solution:}
\begin{itemize}
  \item Verify weather file exists in APSIM Examples folder
  \item If not, download from APSIM repository or use a different weather file
  \item Alternatively, use absolute path (if sharing files, portability is lost)
\end{itemize}

\section{JSON Parsing Errors}

\begin{warningbox}
\textbf{Error message:} ``Error parsing .apsimx file: JSON format invalid''
\end{warningbox}

\textbf{Cause:} Manually edited .apsimx file has JSON syntax error (mismatched braces, quotes, etc.)

\textbf{Solution:}
\begin{itemize}
  \item Use JSON validator (paste file contents at jsonlint.com)
  \item Use APSIM GUI to edit instead of text editor
  \item If editing manually, use code editor with JSON highlighting (VS Code, Sublime, etc.)
\end{itemize}

\section{Simulation Hangs or Crashes}

\begin{warningbox}
\textbf{Symptom:} Simulation starts but doesn't complete; runs for hours.
\end{warningbox}

\textbf{Possible causes:}
\begin{enumerate}
  \item Very long time period (e.g., 1000 years) or complex soil model
  \item Infinite loop in Manager code
  \item Numerical instability (rare)
\end{enumerate}

\textbf{Solution:}
\begin{itemize}
  \item Start with short simulations (1--5 years) to test setup
  \item Check Manager JavaScript for logical errors
  \item If simulation is legitimately long, be patient or increase hardware
\end{itemize}

\section{Getting Help}

If you encounter an error not listed above:

\begin{enumerate}
  \item Check \textbf{APSIM Documentation}: \url{https://www.apsim.info}
  
  \item Post on \textbf{APSIM Forum}: \url{https://github.com/APSIMInitiative/ApsimX/discussions}
  
  \item Consult \textbf{Wiki Resources}: \texttt{FAQ\_Crop\_Won't\_Emerge}, \texttt{FAQ\_Yield\_Unrealistic}, \texttt{FAQ\_Report\_Variables\_Empty}
  
  \item Email instructor with:
    \begin{itemize}
      \item .apsimx file (or key excerpts)
      \item Error message (screenshot)
      \item What you were trying to do (steps)
      \item Expected vs. observed behavior
    \end{itemize}
\end{enumerate}

% ============================================================================
% END APPENDIX B
% ============================================================================
